---
title: 'multitask pattern similarity: high versus low, 2trpk'
author: "michael freund"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: pygments
    self_contained: true
    lightbox: true
    gallery: true
    
---


# intro


supplementary analysis for jocn manuscript


```{r setup, include = FALSE}

source(here::here("code", "_packages.R"))
source(here("code", "_vars.R"))

conditions <- combo_paste(tasks, c("hi", "lo"))
n.dim <- length(conditions)
tris <- fread(here("in", "rsamods.csv")) %>% filter(grepl("run1", .row), grepl("run1", .col))
X <- tris %>% select(where(is.numeric)) %>% as.matrix

rsarray <- readRDS(here("out", "corr-biased_unpre_jocn_2trpk.RDS"))

stats.subjs <- fread(here("out", "subjs_jocn.csv"))

theme_mat <- list(
  theme(axis.text.y = element_text(), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  )


```



## Notes


**Data/sample/preproc**

  * UB55 sans DMCC52*
  * surface + fMRIprep
  * schaefer 400-07
  
**1st-Level GLM**

  * 2trpk, runwise GLMs
  
Target TRs:

  * Axcpt: `r target.trs$Axcpt`
  * Cuedts: `r target.trs$Cuedts`
  * Stern: `r target.trs$Stern`
  * Stroop: `r target.trs$Stroop`

GLM contrasts:

  * Axcpt: $\text{BY} - \text{BX}$
  * Cuedts: $(\text{InConSwitch} + \text{InConRepeat} - \text{ConSwitch} - \text{ConRepeat})/2$
  * Stern: $\text{LL5RN} - \text{LL5NN}$
  * Stroop: $(\text{PC50InCon} + \text{biasInCon} - \text{PC50Con} - \text{biasCon})/2$


**RSA**

Similarity measure:

  * Similarity measure: standard linear correlation
  * NOT cross-valdiated (e.g., not cv-Euclidean/Mahalanobis)
  * NOT prewhitened
  * betas averaged across run, within task, for RSA

models:

  * focused only on between-task correlations
  * atanh transformed


```{r fig.width = 16, echo = FALSE}

## build models

rsm.empty <- matrix(0, ncol = n.dim, nrow = n.dim, dimnames = list(conditions, conditions))
rsm.task <- rsm.empty
rsm.hi <- rsm.empty
rsm.lo <- rsm.empty
rsm.hilo <- rsm.empty

for (task.i in tasks) rsm.task[grepl(task.i, conditions), grepl(task.i, conditions)] <- 1
rsm.hi[grepl("hi", conditions), grepl("hi", conditions)] <- 1
rsm.lo[grepl("lo", conditions), grepl("lo", conditions)] <- 1
rsm.hilo[grepl("lo", conditions), grepl("hi", conditions)] <- 1
rsm.hilo[grepl("hi", conditions), grepl("lo", conditions)] <- 1

## plot


plot_grid(
  plot_grid(
    rsm.task %>% matplot + theme_mat + labs(title = "task (cells excluded)"),
    rsm.hi   %>% matplot + theme_mat + labs(title = "hihi"),
    rsm.lo   %>% matplot + theme_mat + labs(title = "lolo"),
    rsm.hilo   %>% matplot + theme_mat + labs(title = "hilo"),
    nrow = 1
  ),
  rel_heights = c(0.1, 1)
)


```



# quick look: all correlations

```{r fig.width = 10, fig.height = 7, echo = FALSE, message = FALSE}

p.allcor.density <- stats.subjs %>%
  
  ggplot(aes(b, y = term, fill = term)) +
  geom_density_ridges() +
  
  scale_fill_brewer(type = "qual", palette = 5) +
  
  theme(legend.position = "none") +
  labs(x = "cor (z)")


p.allcor.box <- stats.subjs %>%
  
  ggplot(aes(term, b, fill = term)) +
  geom_hline(yintercept = 0) +
  geom_boxplot(width = 0.1, notch = TRUE) +
  
  scale_fill_brewer(type = "qual", palette = 5) +
  
  theme(legend.position = "none") +
  labs(y = "cor (z)") 



p.allcor.network <- stats.subjs %>%
  
  ggplot(aes(b, y = term, fill = term)) +
  geom_density_ridges() +
  
  scale_fill_brewer(type = "qual", palette = 5) +
  
  facet_grid(
    rows = vars(get.network(parcel)), 
    labeller = labeller(run = c(bt = "between-run", wn = "within-run")),
    scales = "free_x"
    ) +
  
  theme(legend.position = "none") +
  labs(y = "cor (z)")


plot_grid(
  cowplot.title("Pattern correlations by components of RSA model"),
  plot_grid(plot_grid(p.allcor.density, p.allcor.box, ncol = 1), p.allcor.network, nrow = 1),
  ncol = 1, 
  rel_heights = c(0.1, 1)
)


```


# Parcel-wise stats

## stats

```{r}

## calculate contrasts between correlations (z values):

contrasts.subjs <- stats.subjs %>%
  
  pivot_wider(names_from = "term", values_from = "b") %>%
  
  mutate(
    hihi_lolo = hihi - lolo,
    hihi_hilo = hihi - hilo,
    lolo_hilo = lolo - hilo
    ) %>%
  
  pivot_longer(
    cols = c("hihi_hilo", "lolo_hilo", "hihi", "lolo", "hilo", "hihi_lolo"), 
    names_to = "contrast", values_to = "b"
    )


## subset DMCC35 ROIs and average correlations over ROI by subject:

contrasts.subjs.35 <- contrasts.subjs %>%
  
  filter(
    parcel %in% parcellation$key[dmcc35], 
    contrast %in% c("hihi", "hilo", "lolo")
  ) %>%

  group_by(subj, contrast) %>%
  summarize(b = mean(b))


## print group statistics

contrasts.subjs.35 %>%
  
  group_by(contrast) %>%
  summarize(
    p = t.test(b)$p.value,
    statistic = t.test(b)$statistic,
    b = mean(b),
    .groups = "drop_last"
    ) %>% 
  kable


```

## plot

```{r}

p.conditional <- contrasts.subjs.35 %>%

  ggplot(aes(contrast, b)) +
  
  geom_hline(yintercept = 0) +
  
  stat_summary(
    fun = mean, geom = "bar", fill = "grey60",
    width = 0.5
    ) +
  stat_summary(
    fun.data = "mean_cl_boot", 
    geom = "errorbar", width = 0, size = 1
    ) +
  
  labs(y = "Mean correlation (z)", x = "Condition pair") +
  scale_x_discrete(labels = c("Cor(Hi, Hi)", "Cor(Hi, Lo)", "Cor(Lo, Lo)")) +
  
  theme_minimal(base_size = 14) +
  
  theme(
   legend.position = "none", 
   panel.grid      = element_blank(), 
   panel.border    = element_blank(),
   axis.line.y     = element_line(),
   axis.ticks.y    = element_line(),
   axis.text.x = element_text(size = rel(0.75))
  ) +
  
  coord_capped_cart(left = "both") +
  
  annotate(geom = "segment", x = 1, xend = 2, y = -0.04, yend = -0.04, size = 0.75) +
  annotate(geom = "segment", x = 1, xend = 3, y = -0.06, yend = -0.06, size = 0.75) +
  annotate(geom = "segment", x = 2, xend = 3, y = -0.08, yend = -0.08, size = 0.75)


p.conditional


ggsave(
  here("code", "fig_crosstasksimil_conditoinal.tiff"),
  p.conditional, device = "tiff", dpi = 600, width = 8, height = 6, unit = "cm"
  )


```


