---
title: "activation profile analysis"
author: "michael freund"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: pygments
    self_contained: true
    lightbox: true
    gallery: true
---
  

# about

supplementary analysis for jocn manuscript

  
```{r setup, message = FALSE, warning = FALSE}

source(here::here("code", "_packages.R"))
source(here("code", "_vars.R"))


## read data:

a <- readRDS(here("in", "roistats_hilo_target_2trpk_unshifted.RDS"))

## get subjs with complete data

is.missing.fmri <- apply(a, 1, function(x) any(is.na(x)))
subjs.missing.fmri <- names(is.missing.fmri)[is.missing.fmri]
subjs.complete <- sort(setdiff(subjs, subjs.missing.fmri))
a <- a[subjs.complete, , , ]


## subset data by DMCC35 parcels

X <- a[, parcellation$key[dmcc35], , ]
dim(X) <- c(length(subjs.complete), length(parcellation$key[dmcc35])*length(tasks), length(sessions))
dimnames(X) <- list(
  subj = subjs.complete, 
  parcel_task = combo_paste(parcellation$key[dmcc35], tasks), 
  session = sessions
  )


## simulate noise (for false positive control analysis)

X_noise <- array(rnorm(prod(dim(X))), dim = c(dim(X)), dimnames = dimnames(X))
 

```


## notes


**Data/sample/preproc**

  * Subjects: `r length(subjs.complete)` included (with complete fMRI data; INCLUDES TWINS)
  * surface + fMRIprep
  * schaefer 400-07
  
**1st-Level GLM**

  * 2trpk, runwise GLMs
  
Target TRs:

  * Axcpt: `r target.trs$Axcpt`
  * Cuedts: `r target.trs$Cuedts`
  * Stern: `r target.trs$Stern`
  * Stroop: `r target.trs$Stroop`

GLM contrasts:

  * Axcpt: $\text{BY} - \text{BX}$
  * Cuedts: $(\text{InConSwitch} + \text{InConRepeat} - \text{ConSwitch} - \text{ConRepeat})/2$
  * Stern: $\text{LL5RN} - \text{LL5NN}$
  * Stroop: $(\text{PC50InCon} + \text{biasInCon} - \text{PC50Con} - \text{biasCon})/2$



# LDA projection


```{r ldaprojection}


lda_proj <- function(data_matrix, whiten) {
  # data_matrix <- X
  
  mu_k <- apply(data_matrix, 2:3, mean)  ## mean profile of each session (over subjs)
  B_k <- cov(t(mu_k))  ## between-session covariance matrix
  
  
  if (whiten == TRUE) {
    
    data_matrix_c <- sweep(data_matrix, 2:3, mu_k, "-")  ## center each session's profiles
    data_matrix_c <- aperm(data_matrix_c, c(1, 3, 2))
    dim(data_matrix_c) <- c(prod(dim(data_matrix_c)[1:2]), dim(data_matrix_c)[3])  ## concatenate sessions along rows
    Winv <- invcov.shrink(data_matrix_c, verbose = FALSE)  ## within-session (noise) inverse covariance matrix
    
  } else {
    
    Winv <- diag(dim(data_matrix)[2])
    
  }
  
  svd_W <- svd(crossprod(Winv, B_k))
  V <- svd_W$u[, 1:2] %*% diag(svd_W$d[1:2]^2)  ## linear discriminants (scaled by variances)
  
  V
  
}



loocv_lda_proj <- function(data_matrix, .whiten) {
  # subj.i <- 1; data_matrix = X
  
  l <- vector("list", length(dim(data_matrix)[1]))
  
  for (subj.i in seq_len(dim(data_matrix)[1])) {
    # subj.i = 1
    
    X_train <- data_matrix[-subj.i, , ]
    X_test <- data_matrix[subj.i, , ]
    
    mu_train <- apply(X_train, 2, mean)  ## training set mean of (session) means
    X_test <- sweep(X_test, 1, mu_train)  ## center test set on mean of means in training set
    
    V_train <- lda_proj(X_train, .whiten)  ## discriminants
    
    X_train_proj <- crossprod(X_test, V_train)
    X_train_proj <- data.table::as.data.table(X_train_proj, keep.rownames = TRUE)
    X_train_proj$subj <- subj.i
    
    l[[subj.i]] <- X_train_proj
    
  }
  
  rbindlist(l)
    
}


boot_lda_loocv <- function(d, n_iter = 1E4, n_cores = n.cores / 2) {
  # d = X_proj_loocv_lda; n_cores = n.cores / 2; n_iter = 1E4
  
  n.subj <- length(unique(d$subj))
  
  set.seed(0)
  
  cl <- makeCluster(n_cores)
  registerDoParallel(cl)
  
  l <- foreach(ii = seq_len(n_iter), .inorder = FALSE, .packages = "data.table") %dorng% {
    
    inds <- sample.int(n.subj, replace = TRUE)
    
    d_ii <- d[subj %in% inds]
    d_ii_bar <- d_ii[, .(V1 = mean(V1), V2 = mean(V2)), by = rn]
    d_ii_bar$iter <- ii
    
    d_ii_bar
    
  }
  
  stopCluster(cl)
  
  rbindlist(l)
  
}


## lda, pca, noise ----


X_proj_loocv_lda <- loocv_lda_proj(X, .whiten = TRUE)
X_bar_loocv_lda <- boot_lda_loocv(X_proj_loocv_lda)

X_proj_loocv_pca <- loocv_lda_proj(X, .whiten = FALSE)
X_bar_loocv_pca <- boot_lda_loocv(X_proj_loocv_pca)

X_proj_loocv_noise <- loocv_lda_proj(X_noise, .whiten = TRUE)
X_bar_loocv_lda_noise <- boot_lda_loocv(X_proj_loocv_noise)


```


## plot


```{r}

## plot ----

p_loocv_lda_points <- X_bar_loocv_lda %>%
  
  ggplot(aes(V1, V2, fill = rn, color = rn)) +
  geom_point(size = 0.1, alpha = 0.05) +
  scale_fill_manual(values = colors.session) +
  scale_color_manual(values = colors.session) +
  
  geom_text(
    data =
      data.frame(
        x = c(-6, -1, 2.25),
        y = c(-0.75, 1, -1),
        # y = c(-1.5, -1.5, -1.5),
        fold = "fold 1",
        label = sessions,
        rn = sessions
      ),
    aes(x = x, y = y, label = label, color = rn), hjust = 0, size = 4, fontface = "bold") +

  theme(legend.position = "none", panel.grid = element_blank(), axis.ticks = element_line()) +
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1)) +

  labs(x = "LD1", y = "LD2", title = "LDA (loocv)")


p_loocv_pca_points <- X_bar_loocv_pca %>%
  
  ggplot(aes(-V1, V2, fill = rn, color = rn)) +
  scale_alpha_continuous(range = c(0, 1)) +

  geom_point(size = 0.1, alpha = 0.05) +
  scale_fill_manual(values = colors.session) +
  scale_color_manual(values = colors.session) +
  
  theme(legend.position = "none", panel.grid = element_blank(), axis.ticks = element_line()) +
  scale_x_continuous(breaks = c(0, 0.01)) +
  scale_y_continuous(breaks = c(0, 0.01)) +
  
  labs(x = "PC1", y = "PC2", title = "PCA (loocv)")


p_noise_lda_points <- X_bar_loocv_lda_noise %>%
  
  ggplot(aes(V1, V2, fill = rn, color = rn)) +
  geom_point(size = 0.1, alpha = 0.05) +
  scale_fill_manual(values = colors.session) +
  scale_color_manual(values = colors.session) +

  theme(legend.position = "none", panel.grid = element_blank(), axis.ticks = element_line()) +
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1)) +

  labs(x = "LD1", y = "LD2", title = "null (simulated)")


p_loocv_points <- grid.arrange(
  p_loocv_pca_points, p_loocv_lda_points, p_noise_lda_points, ncol = 3, 
  top = "resampled session centroids (dims: dmcc35 parcels*task)"
  )


```



# cross-validated mahalanobis


```{r contrasts}


distance_cv <- function(B1, B2, m, regressors) {
  # B1 = B[, , 1]; B2 = B[, , 2]
  
  ## D = diag(A_1 %*% A_2), where A_1 = m %*% B_1 and A_2 = m %*% B_2
  ## i.e., diagonal of product of pattern contrast matrices.
  ## aka, vector of all pairwise distances (includes diag):
  D <- colSums(t(m %*% B1 * m %*% B2))
  matrix(D, ncol = length(regressors), dimnames = list(.row = regressors, .col = regressors))  ## reshape to matrix
  
}



contrast_matrix <- function(n, condition.names) {
  # n <- 10
  # condition.names <- letters[1:n]
  
  if (n < 2) stop("you need more than 1 condition you dummy")
  
  W <- matrix(0, nrow = n^2, ncol = n)
  
  if (missing(condition.names)) {
    dimnames(W) <- list(contrast = NULL, condition = NULL)
  } else {
    dimnames(W) <- list(
      contrast = paste0(rep(condition.names, each = n), "_", rep(condition.names, n)),
      condition = condition.names
    )
  }
  
  for (condition.i in seq_len(n)) {
    # condition.i = 1
    
    row.beg <- (condition.i - 1) * n + 1
    row.end <- (condition.i - 1) * n + n
    W.i <- W[row.beg:row.end, ]  ## square matrix; the contrasts that define a column of the similarity matrix
    
    W.i[, condition.i] <- 1  ## the condition to which all others are contrasted
    diag(W.i) <- diag(W.i) - 1  ## all others
    
    W[row.beg:row.end, ] <- W.i
    
  }
  
  W
  
}


M <- contrast_matrix(3, sessions)  ## matrix that calculates all pairwise contrasts between conditions


## est invcov:

mu_k <- apply(X, 2:3, mean)  ## mean profile of each session (over subjs)
data_matrix_c <- sweep(X, 2:3, mu_k)  ## center each session's profiles
data_matrix_c <- aperm(data_matrix_c, c(1, 3, 2))
dim(data_matrix_c) <- c(prod(dim(data_matrix_c)[1:2]), dim(data_matrix_c)[3])  ## concatenate sessions along rows
Winv <- invcov.shrink(data_matrix_c, verbose = FALSE)

## est cross-validated distances:

D <- array(
  NA, 
  dim = c(length(sessions), length(sessions), dim(X)[1]), 
  dimnames = list(session = sessions, session = sessions, subj = dimnames(X)[[1]])
  )

for (subj.i in seq_len(dim(X)[1])) {
  # subj.i = 1
  
  x_train <- X[subj.i, , ]
  x_test <- X[-subj.i, , ]
  x_test <- apply(x_test, 2:3, mean)
  
  D[, , subj.i] <- distance_cv(t(Winv %*% x_train), t(x_test), M, regressors = sessions)

}

## unwrap to data.table:

l <- vector("list", dim(X)[1])
for (subj.i in seq_along(l)) {
  
  d <- D[, , subj.i]
  v <- d[lower.tri(d)]
  names(v) <- c("bas_pro", "bas_rea", "pro_rea")
  l[[subj.i]] <- as.data.table(v, keep.rownames = TRUE)
  
}
d <- rbindlist(l, idcol = "subj")
```

## plot

```{r}

## stats:

d_stats <- d %>%
  group_by(rn) %>%
  summarize(
    b = mean(v),
    s = t.test(v, alternative = "greater")$statistic,
    p = t.test(v, alternative = "greater")$p.value
    )

d_stats_pairwise <- d %>%
  pivot_wider(names_from = "rn", values_from = "v") %>%
  summarize(
    
    b_bp_br = mean(bas_pro - bas_rea),
    b_br_rp = mean(bas_rea - pro_rea),
    
    t_bp_br = t.test(bas_pro, bas_rea, paired = TRUE)$statistic,
    t_br_rp = t.test(bas_rea, pro_rea, paired = TRUE)$statistic,
    
    p_bp_br = t.test(bas_pro, bas_rea, paired = TRUE)$p.value,
    p_br_rp = t.test(bas_rea, pro_rea, paired = TRUE)$p.value
    
  )

d_stats %>% kable
d_stats_pairwise %>% kable


## plot:

d %>%
  
  ggplot(aes(rn, v)) +

  geom_hline(yintercept = 0) +
  
  geom_line(aes(group = subj), alpha = 1/3) +
  stat_summary(fun = "mean", geom = "bar", width = 0.5) +
  stat_summary(fun.data = "mean_cl_boot", color = "black", size = 2, geom = "errorbar", width = 0.1) +
  
  labs(
    title = "cross-validated distances (dims: dmcc35 parcel*task)",
    x = "contrast",
    y = "group-mean distance"
    )


```




# figure for manuscript


```{r manuscript-fig}


theme_set(theme_bw(base_size = 10))


centroids <- X_bar_loocv_lda %>% group_by(rn) %>% summarize(V1 = mean(V1), V2 = mean(V2))
centroids_w <- centroids %>% pivot_wider(values_from = c("V1", "V2"), names_from = "rn")


p_points_withstats <-

  X_bar_loocv_lda %>%
  
  ggplot(aes(V1, V2, fill = rn, color = rn)) +
  geom_point(size = 0.2, alpha = 0.1, pch = 16) +
  scale_fill_discrete_qualitative(palette = "Dynamic") +
  scale_color_discrete_qualitative(palette = "Dynamic") +
  
  geom_segment(
    data = centroids_w, inherit.aes = FALSE, size = 0.5, alpha = 1, color = "grey50",
    aes(x = V1_baseline, xend = V1_proactive, y = V2_baseline, yend = V2_proactive)
  ) +
  
  geom_segment(
    data = centroids_w, inherit.aes = FALSE, size = 0.5, alpha = 1, color = "grey50",
    aes(x = V1_baseline, xend = V1_reactive, y = V2_baseline, yend = V2_reactive)
  ) +
    
  geom_segment(
    data = centroids_w, inherit.aes = FALSE, size = 0.5, alpha = 1, color = "grey50",
    aes(x = V1_reactive, xend = V1_proactive, y = V2_reactive, yend = V2_proactive), 
  ) +
    
  geom_point(
    data = centroids, 
    size = 1, pch = 21, color = "grey50"
  ) +
  
  annotate(
    x = (centroids_w$V1_baseline + centroids_w$V1_proactive)/2,
    y = (centroids_w$V2_baseline + centroids_w$V2_proactive)/2,
    geom = "text",
    label = paste0(
      "\nt = ", d_stats %>% filter(rn == "bas_pro") %>% pull("s") %>% round(1)
      # "p ", d_stats %>% filter(rn == "bas_pro") %>% pull("p") %>% format.pval
    ),
    size = 2,
    angle = 12, color = "grey50"
  ) +
  
  annotate(
    x = (centroids_w$V1_baseline + centroids_w$V1_reactive)/2,
    y = (centroids_w$V2_baseline + centroids_w$V2_reactive)/2,
    geom = "text",
    label = paste0(
      "t = ", d_stats %>% filter(rn == "bas_rea") %>% pull("s") %>% round(1), "\n"
      # "p ", d_stats %>% filter(rn == "bas_rea") %>% pull("p") %>% format.pval
    ),
    size = 2,
    angle = -25, color = "grey50"
  ) +
  
    
  annotate(
    x = (centroids_w$V1_proactive + centroids_w$V1_reactive)/2,
    y = (centroids_w$V2_proactive + centroids_w$V2_reactive)/2,
    geom = "text",
    label = paste0(
      "t = ", d_stats %>% filter(rn == "pro_rea") %>% pull("s") %>% round(1), "\n"
      # "p ", d_stats %>% filter(rn == "pro_rea") %>% pull("p") %>% format.pval
    ),
    size = 2,
    angle = 55, color = "grey50"
  ) +

  geom_text(
    data =
      data.frame(
        x = c(-12, -1, -8),
        y = c(3, 4.5, -3),
        label = sessions,
        rn = sessions
      ),
    aes(x = x, y = y, label = label, color = rn), hjust = 0, size = 3, fontface = "bold") +

   theme(
     legend.position = "none", 
     panel.grid      = element_blank(), 
     panel.border    = element_blank(),
     axis.line = element_line(),
     axis.ticks = element_line(),
     axis.title = element_blank()
    ) +
    
  coord_capped_cart(left = "both", bottom = "both") +
  scale_x_continuous(breaks = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 1))


p_points_withstats

ggsave(
  here("out", "fig_multivariate_triangle.tiff"),
  p_points_withstats, device = "tiff", dpi = 600, width = 7, height = 6, unit = "cm"
  )


```

